### 2015-04-09 thesis-proposal/problem.tex
rw

%Well-designed general-purpose programming languages give library providers the power to  modularly express a wide variety of useful constructs using a comparatively small set of primitives. %New primitives, and thus new dialects of these languages, should arise very rarely. 
%New primitives (and new \emph{dialects} of the language). %Extending a general-pandurpose language with new primitives, i.e.  forming an extended \emph{dialect} of the language, should, given a sufficiently expressive such language, be exceedingly rare. %Indeed, a diversity of dialects around a language can be taken as evidence against claims of  generality.%, this can be taken as evidence against claims of its generality. %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved. 
%Many languages claim to achieve this ideal. 
%For example, 
%If we measure the generality of a language by how frequently  dialects of the language emerge, 
%all major contemporary languages seem to leave room for improvement. 
%General-purpose programming languages are a dime a. 
%Among the chief goals in the design of a general-purpose programming languages must achieve \emph{stability}: they  very rarely needs to be extended or forked into dialects by the user community surrounding it. Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in. 
Functional programming languages like Standard ML (SML) and Haskell descend directly from elegant typed lambda calculi, diverging mainly in that they strive to present a more pragmatic concrete syntax and type structure to programmers. In other words, their design considers important human factors in addition to fundamental metatheoretic issues. For example, both languages build in record types, generalizing the nullary and binary product types more common in simpler calculi, because labels are cognitively useful.
%Minimal typed lambda calculi play a central role in the study of programming languages, and serve as direct inspiration for  functional programming languages like ML. to be useful in practice, it must present a decidedly more elaborate concrete syntax and type structure to programmers, designed considering various human factors in addition to fundamental metatheoretic issues. For example, while it is quite common to study minimal calculi that build in only nullary and binary product types, Standard ML (SML) builds in record types, because labels are cognitively useful. %syntactic and semantic primitives that capture common idioms naturally and precisely.

Ideally, transitioning from a minimal calculus to a pragmatic programming language would require introducing only a limited number of primitive ``embellishments'' like this. The community around the language would then be able to redirect its efforts exclusively toward the development of a wide variety of useful libraries. Alas, a language design that achieves this ideal has proven elusive, as evidenced by the diverse array of {dialects} that continue to proliferate around all major languages, functional languages included. % For example, let us consider the many dialects of ML. Though they generally agree on the importance of a powerful module system atop a core language based conceptually on the   polymorphic lambda calculus with recursive sum and product types, they all introduce various primitive syntactic and semantic extensions and tweaks of their own design (we cover several examples below). 
%Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
In fact, tools for constructing new ``domain-specific'' dialects seem only to be increasingly popular. 
{This calls for an investigation}: why is it that well-informed programmers and researchers are so often unsatisfied with expressing the constructs that they need in libraries, as modes of use of the ``general-purpose'' functional primitives available today?
%Let us investigate why this is.
%What kinds of constructs are not expressible in languages like SML?

@@

Perhaps the most common  reason may simply be that the \emph{syntactic cost} of expressing a construct of interest using contemporary general-purpose primitives is not always ideal. %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined. 
For example,  we will consider regular expression patterns expressed using abstract data types (in ML, a mode of use of the module system) in Section \ref{sec:syntax}. This  will give  us precisely the semantics that we need, but syntactically, it will leave something to be desired. To avoid this syntactic cost, library providers are frequently tempted to construct a \emph{derived syntactic dialect}, i.e. a dialect specified by context-free elaboration to the existing language, that introduces derived syntax specific to their library. For example, Ur/Web is a  derived syntactic dialect of Ur (itself descended from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes common in web programming \cite{conf/popl/Chlipala15}. In fact, nearly all major programming languages primitively build in derived syntax for constructs that can otherwise be expressed in the standard library, e.g. derived list syntax in SML \cite{harper1997programming,mthm97-for-dart}. Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} (which we will return to later) help lower the engineering costs of constructing such dialects, contributing to their proliferation.% contributing to their proliferation.

@@

More advanced dialects introduce  new type structure, going beyond what is possible by context-free elaboration to the existing language. For example, as  mentioned above, Standard ML goes beyond the nullary and binary product types found in simpler calculi, instead exposing record types. Other dialects have explored  ``record-like'' primitives that support functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
, mutable rows, methods, prototypic dispatch and various other embellishments. Ocaml builds in a  sophisticated object system, polymorphic variants and  conveniences like logic for typechecking operations that use format strings, e.g. $\mathtt{sprintf}$ \cite{ocaml-manual}. ReactiveML builds in primitives for functional reactive programming \cite{mandel2005reactiveml}.  ML5 builds in primitives for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}. Manticore builds in parallel arrays \cite{conf/popl/FluetRRSX07}. Alice ML builds in futures and promises \cite{AliceLookingGlass}. MLj builds in essentially the entire Java programming language, to support typesafe interoperation with existing Java programs \cite{Benton:1999:IWW:317636.317791}. Indeed, perusal of the proceedings of any major programming languages conference will typically yield several more examples. Tools like proof assistants and logical frameworks support specifying and reasoning metatheoretically about dialects like these, and tools like compiler generators and language workbenches lower the implementation cost, again contributing to their increasing proliferation.

@@

The reason why this proliferation of language dialects should be considered alarming is that it is, in an important sense,  anti-modular: a program written in one dialect cannot, in general, safely and idiomatically interface with libraries written in a different dialect. %At best, one can hope that the compilers for the decidedlytwo languages target a common intermediate language,  
To do so would fundamentally require combining the two dialects into a single language. Left unconstrained, this is an ill-posed problem, i.e. given two dialects that are specified by judgements of arbitrary form, it is not clear what it would even mean to combine them. Even when the dialects are specified using a formalism that does provide some way to na\"ively combine two dialects, there is generally no guarantee that the composition will conserve important syntactic and semantic properties. %There is no well-defined mechanism for constructing such a ``combined language'' in general. 
For example, consider two derived syntactic dialects, one building in derived syntax for JSON (a popular data interchange format), the other using a similar syntax for finite mappings of some other type. Though each is known to have an unambiguous concrete syntax in isolation, when their syntax is na\"ively  combined (e.g. by Camlp4), ambiguities arise.  Any putative ``combined language'' must formally be considered a  distinct system for which one must derive all metatheorems of interest anew, guided only informally by those derived for the dialects individually. % would generally attempt to combine their semantics, particularly when they are specified in very different ways. %the ideas housed in one dialect are often only available to programmers willing to do without ideas housed in other dialects. 
%It is thus infeasible to simply allow different contributors to a software system to choose their own favorite dialect for each component they are responsible for. %A complex software system written in multiple distinct language dialects is far too unwieldy for it to be viable. 
%It it clear that dialects are better rhetorical devices than practical engineering artifacts. 
Due to this paucity of modular reasoning principles, language dialects are not practical artifacts for software development ``in the large''. %Large software projects and software ecosystems must pick a single language that does provide powerful modular reasoning principles and, to benefit from them, stay inside it.

Dialects can have an indirect impact on large-scale software development because the language designers that control comparatively popular languages, like Ocaml and Scala, can be convinced to incorporate  ideas from dialects  into backwards compatible language revisions. %These decisions are increasingly influenced by community processes, e.g. the Scala Improvement Process.  %This approach concentrates power as well as responsibility over maintaining metatheoretic guarantees in the hands of a small group of language designers, though increasingly influenced by various community processes (e.g. the Scala Improvement Process). 
%Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
These languages have, in this way, become more expressive over time. However, this expressiveness has come at a cost: they have also ballooned in size. Yet because there appears to be a ``long tail'' of potentially useful primitives (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}),  primitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still often left languishing in ``toy'' dialects. 
Recalling the words of  Reynolds, which are as relevant today as they were nearly half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition. %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.

@@

\newpage %A measured approach to incorporating new primitives into a general-purpose programming language is  sensible because once a primitive is introduced, it becomes entrenched as-is and monopolizes ``syntactic resources'', as we will discuss below. 
This leaves the language design community with two possible paths forward if we wish to keep general-purpose languages small and free of \emph{ad hoc} primitives. One, exemplified (arguably) by SML, is to generally eschew the introduction of new primitives and settle on a set of primitives that sit at a ``sweet spot'' in the overall language design space, accepting that in some circumstances, this trades away expressive power or leads to  high syntactic cost. % that this implies.  
The other path forward is to continue on in the search for even more general language primitives, i.e. primitives that are so expressive that they permit us to degrade a broad class of existing primitives, including those found in dialects, to modularly composable library constructs, where they can be evaluated on their individual merits by programmers and used together without the possibility of conflict. % This should render the construction of new dialects increasingly unnecessary. 
Encouragingly, such new primitives do occasionally arise. For example, Ocaml recently added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}. Using GADTs, Ocaml was able to move some of the machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into the standard library (unfortunately, some machinery remains built in).

Our broad aim is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes even more dramatic steps down this second path. %Similarly, it recently introduced ``open datatypes'', which subsume its previous more specialized exception type, and captures many use cases for .

@@

Verse has a module system taken directly from SML, but its core language is organized in a novel manner around a minimal typed lambda calculus, called the Verse \emph{internal language}, discussed in Section \ref{sec:verse}. In lieu of typical {external} (i.e. programmer-facing) derived syntax and  type structure, Verse introduces two novel primitive constructs:
\begin{itemize}
\item \textbf{Typed syntax macros} (TSMs), which we introduce in Section \ref{sec:syntax}, subsume the need for derived syntax specific to library constructs, e.g. list syntax as built in to SML, by giving library providers static control over the parsing and elaboration of delimited segments of concrete syntax.
\item \textbf{Metamodules}, which we introduce in Section \ref{sec:metamodules}, give library providers static hooks directly into the semantics of the Verse external language, allowing  them to define, for example, the type structure of records (and various ``record-like'' constructs, as discussed above) by type-directed translation to the more minimal internal language (which builds in nullary and binary products, but not records).
\end{itemize} 
Both TSMs and metamodules can be understood as \emph{metaprogramming} primitives, because they involve static code generation. %We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.

The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to give library providers decentralized control over aspects of the language that have typically been under the monolithic control of the language designer. If we are not careful, many of the problems discussed above as inherent to combining distinct language dialects could simply manifest themselves in the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer. The chief characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended. We will avoid this somewhat confusing terminology.} Our main technical contributions will come in showing how to address these problems in a principled manner. In particular, syntactic conflicts will be impossible by construction and  we will validate the code  statically generated by TSMs and metamodules to maintain a hygienic type discipline and, most uniquely, powerful modular reasoning principles -- library providers will be able to reason about the constructs that they have defined in isolation, and library clients will be able to use them safely in any combination, without the possibility of conflict.\footnote{We  assume that simple naming conflicts can be avoided extrinsically, e.g. by using a URI-based naming scheme as in the Java ecosystem.}

@@

\section{Summary of Judgements}\label{sec:verse}
To situate ourselves within a formal framework, let us begin with a brief summary of how Verse is organized and specified. Verse consists of a \emph{module language} atop a \emph{core language}. The module language is based directly on SML's module language,  which we assume working familiarity with for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036}. We will give examples of its use in both Section \ref{sec:syntax} and Section \ref{sec:metamodules}, but do not plan to specify it in detail.

>>>>

%Well-designed general-purpose programming languages give library providers the power to  modularly express a wide variety of useful constructs using a comparatively small set of primitives. %New primitives, and thus new dialects of these languages, should arise very rarely. 
%New primitives (and new \emph{dialects} of the language). %Extending a general-pandurpose language with new primitives, i.e.  forming an extended \emph{dialect} of the language, should, given a sufficiently expressive such language, be exceedingly rare. %Indeed, a diversity of dialects around a language can be taken as evidence against claims of  generality.%, this can be taken as evidence against claims of its generality. %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved. 
%Many languages claim to achieve this ideal. 
%For example, 
%If we measure the generality of a language by how frequently  dialects of the language emerge, 
%all major contemporary languages seem to leave room for improvement. 
%General-purpose programming languages are a dime a. 
%Among the chief goals in the design of a general-purpose programming languages must achieve \emph{stability}: they  very rarely needs to be extended or forked into dialects by the user community surrounding it. Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in. 
Functional programming languages like Standard ML (SML) and Haskell descend directly from mathematically elegant typed lambda calculi, diverging mainly in that they present a more pragmatic concrete syntax and type structure to programmers. In other words, their design considers important human factors in addition to fundamental metatheoretic issues. For example, they build in record types, generalizing the nullary and binary product types more common in simpler calculi, because labels are cognitively useful. Similarly, they build in derived concrete syntax for the list datatype. 
%Minimal typed lambda calculi play a central role in the study of programming languages, and serve as direct inspiration for  functional programming languages like ML. to be useful in practice, it must present a decidedly more elaborate concrete syntax and type structure to programmers, designed considering various human factors in addition to fundamental metatheoretic issues. For example, while it is quite common to study minimal calculi that build in only nullary and binary product types, Standard ML (SML) builds in record types, because labels are cognitively useful. %syntactic and semantic primitives that capture common idioms naturally and precisely.

Ideally, transitioning from a minimal calculus to a pragmatic programming language would require introducing only a limited number of primitive ``embellishments'' like this. The community around the language would then be able to redirect its efforts exclusively toward the development of a wide variety of useful libraries. Alas, a language design that achieves this ideal has proven elusive, as evidenced by the diverse array of {dialects} that continue to proliferate around all major languages, functional languages included. % For example, let us consider the many dialects of ML. Though they generally agree on the importance of a powerful module system atop a core language based conceptually on the   polymorphic lambda calculus with recursive sum and product types, they all introduce various primitive syntactic and semantic extensions and tweaks of their own design (we cover several examples below). 
%Tools for  constructing new ``domain-specific'' dialects also continue to proliferate. % The situation is similar across all major language lineages. 
In fact, tools for constructing new ``domain-specific'' dialects seem only to be increasingly popular. 
{This calls for an investigation}: why is it that well-informed programmers and researchers are so often unsatisfied with expressing the constructs that they need in libraries, as modes of use of the ``general-purpose'' functional primitives already available today?
%Let us investigate why this is.
%What kinds of constructs are not expressible in languages like SML?

@@

Perhaps the most common  reason may simply be that the \emph{syntactic cost} of expressing a construct of interest using contemporary general-purpose primitives is not always ideal. %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined. 
For example,  we will consider regular expression patterns expressed using abstract data types (in ML, a mode of use of the module system) in Section \ref{sec:syntax}. This  will give  us precisely the semantics that we seek to express, but syntactically, the usual approach of using strings to introduce patterns actually leaves much to be desired. In response, we may be tempted to construct a \emph{derived syntactic dialect}, i.e. a dialect specified by context-free elaboration to the existing language, that introduces derived syntax specific to our library, just as languages like SML have introduced derived syntax for constructs in the standard library, like lists. This is not at all an isolated example. For example, Ur/Web is a  derived syntactic dialect of Ur (itself descended from ML) that builds in derived syntax for SQL queries, HTML elements and other datatypes common in web programming \cite{conf/popl/Chlipala15}. %In fact, nearly all major programming languages primitively build in derived syntax for constructs that can otherwise be expressed in the standard library, e.g. derived list syntax in SML \cite{harper1997programming,mthm97-for-dart}. 
Tools like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework} (which we will return to later) are designed to lower the engineering costs of constructing such dialects, contributing to their proliferation.% contributing to their proliferation.

@@

More advanced dialects introduce  new type structure, going beyond what is possible by context-free elaboration to the existing language. For example, as  mentioned above, Standard ML goes beyond the nullary and binary product types found in simpler calculi, instead exposing record types. Other dialects have explored  ``record-like'' primitives that support functional update and extension operators, width and depth coercions (sometimes implicit)%\cite{Cardelli:1984:SMI:1096.1098}
, mutable rows, methods, prototypic dispatch and various other embellishments. Ocaml builds in a  sophisticated object system, polymorphic variants and  conveniences like logic for typechecking operations that use format strings, e.g. $\mathtt{sprintf}$ \cite{ocaml-manual}. ReactiveML builds in primitives for functional reactive programming \cite{mandel2005reactiveml}.  ML5 builds in primitives for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}. Manticore builds in parallel arrays \cite{conf/popl/FluetRRSX07}. Alice ML builds in futures and promises \cite{AliceLookingGlass}. MLj builds in essentially the entire Java programming language, to support typesafe interoperation with existing Java programs \cite{Benton:1999:IWW:317636.317791}. Indeed, perusal of the proceedings of any major programming languages conference will typically yield several more examples. Tools like proof assistants and logical frameworks support specifying and reasoning metatheoretically about dialects like these, and tools like compiler generators and language workbenches lower their implementation cost, again contributing to their continued proliferation.

@@

The reason why this proliferation of language dialects should be considered alarming is that it is, in an important sense,  anti-modular: a program written in one dialect cannot, in general, safely and idiomatically interface with libraries written in a different dialect. %At best, one can hope that the compilers for the decidedlytwo languages target a common intermediate language,  
To do so would fundamentally require combining the two dialects into a single language. Left unconstrained, this is an ill-posed problem, i.e. given two dialects that are specified by judgements of arbitrary form, it is not clear what it would even mean to combine them. Even when the dialects are specified using some formalism that does provide a means to na\"ively combine two dialects, there is generally no guarantee that the composition will conserve important syntactic and semantic properties. %There is no well-defined mechanism for constructing such a ``combined language'' in general. 
For example, consider two derived syntactic dialects, one building in derived syntax for JSON (a popular data interchange format), the other using a similar syntax for finite mappings of some other type. Though each is known to have an unambiguous concrete syntax in isolation, when their syntax is na\"ively  combined (e.g. by Camlp4), ambiguities arise.  Any putative ``combined language'' must formally be considered a  distinct system for which one must derive all metatheorems of interest anew, guided only informally by those derived for the dialects individually. % would generally attempt to combine their semantics, particularly when they are specified in very different ways. %the ideas housed in one dialect are often only available to programmers willing to do without ideas housed in other dialects. 
%It is thus infeasible to simply allow different contributors to a software system to choose their own favorite dialect for each component they are responsible for. %A complex software system written in multiple distinct language dialects is far too unwieldy for it to be viable. 
%It it clear that dialects are better rhetorical devices than practical engineering artifacts. 
Due to this paucity of modular reasoning principles, language dialects are not practical artifacts for software development ``in the large''. %Large software projects and software ecosystems must pick a single language that does provide powerful modular reasoning principles and, to benefit from them, stay inside it.

Dialects do have an indirect impact on large-scale software development because the language designers that control comparatively popular languages, like Ocaml and Scala, can be convinced to incorporate  ideas from dialects  into backwards compatible language revisions. %These decisions are increasingly influenced by community processes, e.g. the Scala Improvement Process.  %This approach concentrates power as well as responsibility over maintaining metatheoretic guarantees in the hands of a small group of language designers, though increasingly influenced by various community processes (e.g. the Scala Improvement Process). 
%Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
In this this way, these languages have become more general over time. However, this  has come at a cost: they have also ballooned in size. And yet, because there appears to be a ``long tail'' of potentially useful primitives (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}),  primitives that are only situationally useful, or that trigger aesthetic disagreements between different factions of the community, are still often left languishing in ``toy'' dialects. 
Recalling the words of  Reynolds, which are as relevant today as they were nearly half a century ago \cite{Reynolds70}:%Because there is no data about how useful a construct is in practice until it is included in a language like this, decisions about which constructs to include are often informed by little more than intuition. %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.

@@

\newpage %A measured approach to incorporating new primitives into a general-purpose programming language is  sensible because once a primitive is introduced, it becomes entrenched as-is and monopolizes ``syntactic resources'', as we will discuss below. 
This leaves the language design community with two possible paths forward if we wish to keep general-purpose languages small and free of \emph{ad hoc} primitives. One, exemplified (arguably) by SML, is to generally eschew the introduction of new primitives and settle on a set of primitives that sit at a ``sweet spot'' in the overall language design space, accepting that in some circumstances, this trades away expressive power or leads to  high syntactic cost. % that this implies.  
The other path forward is to continue on in the search for even more general language primitives, i.e. primitives that are so expressive that they permit us to degrade a broad class of existing primitives, including those found in dialects, to modularly composable library constructs, so that they can be evaluated on their individual merits by programmers and used together without the possibility of conflict. % This should render the construction of new dialects increasingly unnecessary. 
Encouragingly, primitives that have this effect on a language do occasionally arise. For example, Ocaml in a recent revision added support for  ``generalized algebraic data types'' (GADTs), based on research on guarded recursive datatype constructors \cite{XiCheChe03}. Using GADTs, Ocaml was able to move some of the machinery for typechecking operations that use format strings, like \texttt{sprintf}, out of the language and into the standard library (unfortunately, some machinery remains built in). 
Our broad aim is to introduce a new functional programming language called Verse\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work being proposed here, because Wyvern is a group effort evolving independently in some important ways.} that takes even more dramatic steps down this second path. %Similarly, it recently introduced ``open datatypes'', which subsume its previous more specialized exception type, and captures many use cases for .

@@

Verse has a module system taken directly from SML, but its core language is organized in a novel manner around a minimal typed lambda calculus, called the Verse \emph{internal language}, discussed in Section \ref{sec:verse}. In lieu of typical {external} (i.e. programmer-facing) derived syntax and  type structure, Verse introduces two novel primitive constructs:
\begin{itemize}
\item \textbf{Typed syntax macros} (TSMs), which we introduce in Section \ref{sec:syntax}, subsume the need for derived syntax specific to library constructs, e.g. list syntax as built in to SML, by giving library providers static control over the parsing and elaboration of delimited segments of concrete syntax.
\item \textbf{Metamodules}, which we introduce in Section \ref{sec:metamodules}, give library providers static hooks directly into the type-directed translation protocol that governs the Verse external language. For example, library providers will be able to define a metamodule that introduces extensible labeled product types and their associated operators, defining their semantics by type-directed translation to the internal language, which builds in only nullary and binary product types.
\end{itemize} 
Both TSMs and metamodules can be understood as \emph{metaprogramming} primitives, because they involve static code generation. %We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.

The key challenge in the design of these primitives will come in ensuring that they are metatheoretically well-behaved, given that they aim to give library providers decentralized control over aspects of the language that have typically been under the monolithic control of the language designer. If we are not careful, many of the problems discussed earlier as inherent to combining distinct language dialects could simply manifest themselves in the semantics of these primitives.\footnote{This is why languages  like Verse are often called \emph{extensible languages}, though this is somewhat of a misnomer. The chief characteristic of an extensible language is that it \emph{doesn't} need to be extended in situations where other languages would need to be extended. We will avoid this somewhat confusing terminology.} Our main technical contributions will come in showing how to address these problems in a principled manner. In particular, syntactic conflicts will be impossible by construction and  we will validate the code  statically generated by TSMs and metamodules to maintain a hygienic type discipline and, most uniquely, powerful modular reasoning principles. In other words, library providers will have the ability to reason about the constructs that they have defined in isolation, and library clients will be able to use them safely in any combination, without the possibility of conflict.\footnote{We  assume that simple naming conflicts can be avoided extrinsically, e.g. by using a URI-based naming scheme as in the Java ecosystem.}

@@

\section{Summary of Judgements}\label{sec:verse}
To situate ourselves within a formal framework, let us begin with a brief summary of how Verse is organized and specified. Verse consists of a \emph{module language} atop a \emph{core language}. The module language is based directly on SML's module language,  which we assume working familiarity with for the purposes of this proposal \cite{harper1997programming,MacQueen:1984:MSM:800055.802036}. We will give examples of its use in both Section \ref{sec:syntax} and Section \ref{sec:metamodules}, but do not plan to specify it in detail in the thesis.\todo{reminder that we are not going to re-specify the module system}
